Continuous estimates of heat emission at Mt. Ruapehu using the Unscented Kalman Smoother

### 2021-08-02 manuscript/manuscript.tex
version sent to Steve and Tony

%%% Put your definitions there:
\startlocaldefs
\newcommand{\bm}[1]{{\bf #1}}
\endlocaldefs

%%% Begin ...
\begin{document}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Introduction}

@@

Crater Lake (Te Wai A-Moe) on Mt. Ruapehu, an active andesitic stratovolcano in
the center of New Zealand's North Island (Figure \ref{overview}), is one such
example with records of observations dating back to the 19th century
\citep[e.g.][]{Friedlander1898}. Mt. Ruapehu erupted at least 602 times since
1830 with many phreatic and phreato-magmatic eruptions and two major magmatic
episodes \citep{Scott2013}. Although the latest magmatic event between June
1995 to November 1997 and a subsequent dam collapse changed the shape and
volume of the lake, it has been largely unchanged since then and the lake's
bathymetry is well known from times when it was empty [Tony Hurst, personal
communication].

Increases in heat input may be caused by the ascent of magma or
changes in the permeability of the hydrothermal system and are, therefore,
important parameters to be monitored. First attempts to quantify the amount of
heat entering the lake date back to \citep{Dibble1966}. Since 2016
water temperature and lake level have been monitored continuously
and water samples for ion concentration measurements have been taken roughly
every two to three months. This enables us to continuously estimate heat input
into the lake and, in principle, automatically detect changes as soon as they occur.

Inference of the heat input is typically done through a mass and
energy balance calculation: the sum of mass and energy flow into and out of the
lake has to account for the observed changes in lake temperature, water level
and ion dilution \citep[e.g.,][]{Hurst1981, Hurst1991, Stevenson1992,
Fournier2009, Scott1994}. At Crater Lake, and probably at most other volcanic
lakes, this is an underdetermined problem, that is the number of unknowns
exceeds the number of constraints. Hence, it is necessary to constrain some of
the parameters in the mass and energy balance model using prior knowledge. Further,
some observations and input parameters come with large uncertainties and are sampled
irregularly.

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/overview.png}  
\caption{Image of Mt. Ruapehu's crater lake from 2008 (\textcopyright Karen
	Britten, GNS Science). The red O marks the location of the lake outlet. The
	location of Mt. Ruapehu is shown as a red triangle on the insert map.}
  \label{overview}
\end{figure}

\section*{Probabilistic inference method}\label{Pim}

\subsection*{Physical model}\label{phm}

@@

where $T$ is temperature and $c_p$ is the specific heat. For all following
equations we take $H(T_0=0)$ as our reference enthalpy and $c_p=c_w=constant$,
with $c_w$ the specific heat of water, such that the enthalpy of lake water at
temperature $T$ can be written as $H(T) = c_wMT$. Because Ruapehu Crater Lake is
 at an altitude of $\sim$~2500~m and big parts of its catchment are covered by snow
and ice all year round we assume the temperature of incoming water to be at 0
$^{\circ}C$.

@@

where $T$ is the lake water temperature, $\dot{Q_r}$ is
the rate of energy gain due to solar irradiation (short-wavelength radiation),
and $c_wT\dot{M_o}$ is the rate of energy loss through outflow/seepage. $\dot{Q_e}$ is
the rate of surface energy losses due to evaporation (forced and free
convection), sensible heat, and long-wavelength radiation (net loss). Several
empirical and theoretical equations exist to describe the surface losses and
both \cite{Stevenson1992} and \cite{hurstCraterLakeEnergy2015} discuss them in detail. We
follow here mostly the choice of \cite{hurstCraterLakeEnergy2015} as it includes the recent
development by \cite{sartoriCriticalReviewEquations2000} on forced (i.e. wind-induced) convection
(see Appendix \ref{A} for more details). $\dot{Q_{eff}}$ is the effective rate of
energy gain from volcanic sources after heating incoming water:
\begin{equation}
	\dot{Q_{eff}}=\dot{Q_i} - c_wT\dot{M_s}
\end{equation}
with $\dot{Q_i}$ the rate of energy input from volcanic sources.

@@

\begin{figure}
    \includegraphics[width=\textwidth]{figures/sensitivity_analysis.png}  
    \caption{Sensitivity analysis of $\dot{T}$ with respect to $\dot{Q_i}$ 
             and $\dot{M_o}$ (top panel), $\dot{Q_i}$ and season 
             (center panel), and $T$ and windspeed (bottom panel). The model
              parameters that have been held fixed are described in each 
              panel's title. Warmer colors indicate a temperature increase
              ($\dot{T} > 0$) and cooler colors a temperature decrease
              ($\dot{T} < 0$)}
      \label{sens}
\end{figure}

\subsection*{Inference}\label{inf}

The objective of the inference problem is to estimate the rate at which heat
is entering the lake from the hydrothermal system below.
To constrain this we use the currently available observations of $T$, $M$,
$X$, and $\dot{M_o}$ as well as numerical weather model data on windspeed.
Putting it more concisely, we are looking for

@@

The following equations are based on \cite{sarkkaBayesianFilteringSmoothing2013}
which we refer the reader to for more proofs and detailed explanations.

Equation \ref{bsm} is the marginal probability of $\bm{x}_k$ given all
observations until time $t$. We assume that the states of the lake form a
Markov chain, meaning that the state of the lake at time step $k$, $\bm{x}_k$,
only depends on the state at adjacent time steps:

@@

With this assumption and using \textit{Bayes' rule}, Equation \ref{bsm} can be
rewritten as:

\begin{equation}\label{bsm1}
p(\bm{x}_k|\bm{y}_{1:T}) = p(\bm{x}_k|\bm{y}_{1:k})
    \int\left[\frac{p(\bm{x}_{k+1}|\bm{x}_k)p(\bm{x}_{k+1}|\bm{y}_{1:T})}{p(\bm{x}_{k+1}|\bm{y}_{1:k})}\right]d\bm{x}_{k+1}
\end{equation}

@@

using the \textit{Chapman-Kolmogorov} equation we can write:

@@

where $f(\bm{x}_{k-1})$ is the physical model described in Section \ref{phm}
which is strongly non-linear as shown in Figure \ref{sens}. If we assume the errors, $\bm{r}_k$ and
$\bm{q}_{k-1}$, to be normally distributed ($\bm{q}_{k-1} \sim N(\bm{0},
\bm{Q}_{k-1})$, $\bm{r}_k \sim N(\bm{0}, \bm{R}_k)$), we can compute an
approximate solution to \ref{bsm1} using the \textit{Unscented Kalman Filter}
\citep{Merwe2004}.

We estimate $\bm{R}_k$ from the daily variations in the observations but we do
not know $\bm{x}_1$ and $\bm{Q}_{k-1}$. Setting $\bm{x}_1$ to an arbitrary value
within physically reasonable bounds but large uncertainty works well in practice.
While $\bm{Q}_{k}$ can be formally optimised we treat it here as a design
parameter that we modify such that the observations are matched within their
error bounds and the inferred heat input rate varies smoothly. The latter is
motivated by the fact that the physical model described in \ref{phm} is not
able to model short-term transient changes in lake volume and temperature.
These can be caused by rain or increased meltwater influx and the assumption
of a perfectly mixed lake typically does not apply in these situations.

\section*{Results}
\subsection*{Synthetic test}\label{syn_test}
To evaluate the inference scheme we design a synthetic test, consisting of a very
large cylinder, the same volume as Ruapehu Crater Lake. We keep the windspeed
and surface water inflow constant at 4.5 m/s and 10 kt/day, respectively. To
model the outflow we assume the outlet to be a pipe 0.2 $m^2$ in cross section
and then use Bernoulli's equation to model the water outflow based on the water
level in the cylinder above the pipe. We assume that the incoming heat follows
a combination of a third-order polynomial and a sinusoid function and that the
enthalpy of the incoming steam is 2.8 which corresponds to the enthalpy of
saturated steam at hydrostatic pressures similar to those at the bottom of the
lake \cite{Mayhew1978}. To the resulting synthetic observations we add
uncertainties similar to those estimated from real observations (see Section
\ref{realdata}).

@@

\begin{figure}
	\includegraphics[width=\textwidth]{figures/synthetic_inversion_uks.png}  
    \caption{Inversion result for the synthetic test described in Section 
        \ref{syn_test}. Turquoise lines and markers represent hypothetical
        observations; red dashed lines are the inversion results and black
        dashed lines are the true input to the synthetic test.}
  \label{syn_example}
\end{figure}

\subsection*{Real data}\label{realdata}
Temperature at RCL is measured by a integrated-circuit temperature device
(Texas Instruments LM35) located roughly 1.9 m below the lake's current
overflow level. The water level sensor is a hydrostatic pressure sensor (First
Sensor KTE/KTU/KTW6000\ldots CS Series) and it is co-located with the
temperature sensor. Both sensors have a sampling interval of 15 minutes and
transmit data several times a day via satellite to GNS Science, which monitors
New Zealand's geohazards through its GeoNet project (www.geonet.org.nz
 last accessed 16/07/2021).
Water samples are taken 8-10 times a year from which subsequently, among many
other analytes, concentrations of $Mg^{++}$ and $Cl^-$ are determined at GNS
Science's Geothermal Analytical Laboratory. We will focus here on $Mg^{++}$
concentration as it shows less deviation from the assumption of constant
dilution than $Cl^-$.

To estimate uncertainties of temperature and water level observations, we 
compute their daily averages and standard deviations. Uncertainties of $Mg^{++}$
concentrations are estimated directly when two samples were taken on the same day.
Otherwise, their uncertainty is taken to be the median of directly estimated
uncertainties. $\dot{M_o}$ is measured once or twice per year and we further know
from field observations that the lake is not overflowing beneath a certain water
level.

As there is no weather station close to RCL we assume the wind speed from the
New Zealand Metservice's numerical weather prediction model with a standard
deviation of ??.

\begin{figure}
% 	\includegraphics[width=.9\textwidth]{clemb_inference_result}  
% 	\caption{Inference results for the energy input rate (A),
% 		the steam input rate (B), the water inflow (C) and 
% 		outflow rate (D) for temperature, water level and 
% 		$Mg^{++}$ observations at RCL in 2018.
% 		The color shows an interpolation of 1D kernel density 
% 		estimates (kdes) of the posterior samples for each day's 
% 		inference result. The solid black line marks the expected
% 		energy input rate. (E) shows the Bhattacharyya distance 
% 	        between the probability distributions of the observations
% 	        and the probability distributions of the inference results.}
\label{inference_result}
\end{figure}

\section*{Discussion and conclusion}
Although we use a tighter constraint on the
outflow than just the dilution of $Mg^{++}$, inflow and outflow are still
poorly constraint, especially when the lake level is above overflow level. As we already
showed in the synthetic test in Section \ref{syn_test}, the $MG^{++}$ alone has
too large uncertainties to properly constrain outflow rate. The probabilistic
inference results thus help us determine future priorities for field
observations and work is underway to establish a quantitative relationship
between lake level and outflow rate.

\cite{hurstCraterLakeEnergy2015} and \cite{Stevenson1991} both found that they required very
high values of enthalpy for the incoming steam to avoid violating physical
boundary conditions such as negative values for the inflow rate of meteoric
water. Their explanation was the existence of a heat pipe beneath the lake where
gas and fluid phases coexist. As the rising gas enters the lake relatively cool
lake water enters the vent leading to an overall lower addition of mass to the
lake and resulting in an enthalpy value that is double than expected for
saturated steam at hydrostatic pressures likely to be found at the bottom of the
lake. By including uncertainties in our observations we no longer require the
high enthalpy values to satisfy the physical constraints. In fact, since we \textit{a
priori} only allow physically reasonable parameter ranges we find the
combinations of parameter values that reproduce the observations within their
respective uncertainties.

Inferring the heat input rate into a crater lake is a corner piece in estimating the
total energy budget of a volcanic system as, for example, described by \cite{Adams1989a}
for Po\'as volcano in Costa Rica. Similar to a volcano's gas budget, it can either
indicate migration of magma or changes in permeability within the vent system. To
answer whether it is one or the other requires more detailed numerical modelling
of the magmatic and hydrothermal system. Many studies have demonstrated the 
value of non-linear Kalman Filters for data assimilation with very complex numerical 
models \citep{}. The next logical step, therefore, is to include models of magma
degassing and hydrothermal fluid flow to model the heat input rate, volume, and
enthalpy of the steam entering the lake. Because Kalman Filters return a probabilistic
estimate of model parameters, even crude, first-order models could be used
as long as appropriate model uncertainties are assigned.

* probabilistic approach allows for Bayesian comparison between evaporation models

Although an increased heat input rate at Ruapehu
often goes along with an increase in seismic tremor levels (see Figure ??), this is
not always the case and so it is an independent indicator of potential changes in
the volcano.

Figure ?? shows the comparison of daily averages of volcanic tremor amplitudes
(RSAM; reference), heat input rate and water temperature. We can see that heat
input rate correlates well with tremor values with a correlation coefficient of
?. If we only consider heat pulses above ? MW (see dashed line in Figure ?), the
correlation coefficient increases to ?. Although correlation does not imply
causality it would seem plausible that high heat, and therefore gas, fluxes
could generate volcanic tremor. A mechanism of heat flow generating tremor
has been described by \citet{Iwamura2005}.

In conclusion, the combination of the mass and energy model of \citet{Hurst2015}
and \citet{Stevenson1992} combined with the probabilistic inference method we
developed here provides a powerful method to continuously infer the heat input
rate into Ruapehu Crater Lake. It allows us to combine disparate data streams,
quantify our beliefs in the results and thereby augment our monitoring
capabilities.

@@

\begin{tabular}{cp{8cm}}
    A & is the surface of the lake (in $m^2$) \\
    $\sigma$ & is the Stefan-Boltzmann coefficient ($5.67e^{-8} \frac{W}{m^2 K^4}$) \\
    $\epsilon_w$ & is the lake emissivity \\
    $T_s$ & is the water surface temperature (here assumed to be 1 $^{\circ}C$ less than the
    measured temperature) \\
    $\epsilon_a$ & is the effective atmospheric emissivity \\
    $T_a$ & is the effective air temperature (here assumed to be 0.9 $^{\circ}C$) \\ 
\end{tabular}

\subsection{Evaporation}
As proposed by \citet{Hurst2015} we use the equation of \citet{Adams1990} but
replace the term for forced convection by that of \citet{Sartori2000}:

@@

\begin{tabular}{cp{8cm}}
    $T_{sv}$, $T_{av}$ & are the virtual surface and air temperatures respectively.
    This correction accounts for the extra buoyancy of water vapour compared to air
    (see Equation \ref{vtemp} for details) \\
    $e_s$, $e_a$ & are the saturation vapour pressure and the ambient air vapour
    pressure respectively \\
    $L$ & is the latent heat of evaporation \\
    $u$ & is the wind speed \\
    $X_0$ & is the fetch, or characteristic length, of the lake\\
    $P_a$ & is the atmospheric pressure \\
\end{tabular}

@@

\subsection{Sensible heat}
This describes the heat loss due to conduction and convection above a hot lake and according to
\citet{Stevenson1992} the ratio between sensible heat, $Q_{sh}$, and heat loss due
to evaporation, $Q_{evap}$ can be written as:

@@

\begin{tabular}{cp{8cm}}
    $\rho$ & is the air density \\
    $c_a$ & is the specific heat of air\\
\end{tabular}

>>>>

%%% Put your definitions there:
\startlocaldefs
\newcommand{\bm}[1]{{\textbf{#1}}}
\endlocaldefs

%\newcommand{\bm}[1]{{\textbf{#1}}}
%%% Begin ...
\begin{document}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section{Introduction}

@@

Ruapehu Crater Lake (Te Wai A-Moe in Te Reo Maori and referred to as RCL from
hereon) on Mt. Ruapehu, an active andesitic stratovolcano in the center of New
Zealand's North Island (Figure \ref{overview}), is one such example with records
of observations dating back to the 19th century \cite{Friedlander1898}. Mt.
Ruapehu erupted at least 602 times since 1830 with many phreatic and
phreato-magmatic eruptions and two major magmatic episodes \citep{Scott2013}.
Although the latest magmatic event between June 1995 to November 1997 and a
subsequent dam collapse changed the shape and volume of the lake, it has been
largely unchanged since then and the lake's bathymetry is well known from times
when it was empty [Tony Hurst, personal communication].

First attempts to quantify the amount of heat entering the lake date back to
1966 \citep{Dibble1966}. Since 2016 water temperature and lake level have been
monitored continuously and water samples for ion concentration measurements have
been taken roughly every two to three months. This enables us to continuously
estimate heat input into the lake and, in principle, automatically detect
changes as soon as they occur.

Inference of the heat input is typically done through a mass and energy balance
calculation: the sum of mass and energy flow into and out of the lake has to
account for the observed changes in lake temperature, water level and ion
dilution \cite{Hurst1981, Hurst1991, Stevenson1992, Fournier2009, Scott1994}. At
RCL, and probably at most other volcanic lakes, this is an underdetermined
problem, that is the number of unknowns exceeds the number of constraints.
Hence, it is necessary to constrain some of the parameters in the mass and
energy balance model using prior knowledge. Further, some observations and input
parameters come with large uncertainties and are sampled irregularly.

@@

\begin{figure}
\includegraphics[width=12 cm]{figures/overview.png}  
\caption{Image of Mt. Ruapehu's crater lake from 2008 (\textcopyright Karen
	Britten, GNS Science). The red O marks the location of the lake outlet. The
	location of Mt. Ruapehu is shown as a red triangle on the insert map.}
  \label{overview}
\end{figure}

\section{Probabilistic inference method}\label{Pim}

\subsection{Physical model}\label{phm}

@@

where $T$ is temperature and $c_p$ is the specific heat. For all following
equations we take $H(T_0=0~^{\circ}C)$ as our reference enthalpy and
$c_p=c_w=constant$, with $c_w$ the specific heat of water, such that the
enthalpy of lake water at temperature $T$ can be written as $H(T) = c_wMT$.
Because RCL is at an altitude of $\sim$~2500~m and big parts of its catchment
are covered by snow and ice all year round we assume the temperature of incoming
water to be at 0 $^{\circ}C$.

@@

where $T$ is the lake water temperature, $\dot{Q_r}$ is the rate of energy gain
due to solar irradiation (short-wavelength radiation), and $c_wT\dot{M_o}$ is
the rate of energy loss through outflow/seepage. $\dot{Q_e}$ is the rate of
surface energy losses due to evaporation (forced and free convection), sensible
heat, and long-wavelength radiation (net loss). Several empirical and
theoretical equations exist to describe the surface losses and we refer the
reader to previous publications for more detailed discussions
\cite{Stevenson1992, hurstCraterLakeEnergy2015}. There is no consensus on the
best quantitative model of evaporation from warm lakes and, in fact, the most
suited equation may depend on the location of and available observations from
the lake. Appendix \ref{A} describes the equations used in this study in detail
which include the recent development on forced (i.e. wind-induced) convection
\cite{sartoriCriticalReviewEquations2000}.

$\dot{Q_{eff}}$ is the effective rate of energy gain from volcanic sources after
heating incoming water:
\begin{equation}
	\dot{Q_{eff}}=\dot{Q_i} - c_wT\dot{M_s}
\end{equation}
with $\dot{Q_i}$ the rate of energy input from volcanic sources.

@@

\begin{figure}
    \includegraphics[width=12 cm]{figures/sensitivity_analysis.png}  
    \caption{Sensitivity analysis of $\dot{T}$ with respect to $\dot{Q_i}$ 
             and $\dot{M_o}$ (top panel), $\dot{Q_i}$ and season 
             (center panel), and $T$ and windspeed (bottom panel). The model
              parameters that have been held fixed are described in each 
              panel's title. Warmer colors indicate a temperature increase
              ($\dot{T} > 0$) and cooler colors a temperature decrease
              ($\dot{T} < 0$)}
      \label{sens}
\end{figure}

\subsection{Inference}\label{inf}

The objective of the inference problem is to estimate the rate at which heat
is entering the lake from the hydrothermal system below ($\dot{Q_i}$).
To constrain this we use the currently available observations of $T$, $M$,
$X$, and $\dot{M_o}$. Putting it more concisely, we are looking for

@@

Solving Equation \ref{bsm} using standard strategies like Markov Chain Monte
Carlo is computationally very expensive as the number of variables grows
exponentially with the length of the timeseries. An alternative are Bayesian
Filtering and Smoothing algorithms of which the Kalman Filter and Smoother are
the closed form solution for linear models \cite{Kalman1960, Rauch1965,
sarkkaBayesianFilteringSmoothing2013}.

Equation \ref{bsm} is the conditional probability of $\bm{x}_k$ given all
observations until time $t$. We assume that the states of the lake form a
Markov chain, meaning that the state of the lake at time step $k$, $\bm{x}_k$,
only depends on the state at adjacent time steps:

@@

With this assumption and using \textit{Bayes' rule}, Equation \ref{bsm} can be
rewritten \cite{sarkkaBayesianFilteringSmoothing2013} as:

\begin{equation}\label{bsm1}
p(\bm{x}_k|\bm{y}_{1:t}) = p(\bm{x}_k|\bm{y}_{1:k})
    \int\left[\frac{p(\bm{x}_{k+1}|\bm{x}_k)p(\bm{x}_{k+1}|\bm{y}_{1:t})}{p(\bm{x}_{k+1}|\bm{y}_{1:k})}\right]d\bm{x}_{k+1}
\end{equation}

@@

using the \textit{Chapman-Kolmogorov} equation
\cite{sarkkaBayesianFilteringSmoothing2013} we can write:

@@

where $f(\bm{x}_{k-1})$ is the physical model described in Section \ref{phm}
propagated forward in time using a fourth-order Runge-Kutta scheme. If we 
assume the errors, $\bm{r}_k$ and $\bm{q}_{k}$, to be normally distributed
($\bm{q}_{k} \sim N(\bm{0},
\bm{Q}_{k})$, $\bm{r}_k \sim N(\bm{0}, \bm{R}_k)$), we can compute an
approximate solution to \ref{bsm1} using the \textit{Unscented Kalman Filter}
\citep{Merwe2004, sarkkaBayesianFilteringSmoothing2013}.

We estimate $\bm{R}_k$ from the daily variations in the observations but we do
not know $\bm{x}_1$ and $\bm{Q}_{k}$. Setting $\bm{x}_1$ to an arbitrary value
within physically reasonable bounds but large uncertainty works well in
practice. While $\bm{Q}_{k}$ can be formally optimised we treat it here as a
design parameter that we modify such that the observations are matched within
their error bounds and the inferred heat input rate varies smoothly. The latter
is motivated by the fact that the physical model described in \ref{phm} is not
able to model short-term transient changes in lake volume and temperature. These
can be caused by rain or increased meltwater influx and the underlying
assumption of a perfectly mixed lake typically does not apply in these
situations.

\section{Results}
\subsection{Synthetic test}\label{syn_test}

To evaluate the inference scheme we design a synthetic test, consisting of a
very large cylinder, the same volume as RCL. We keep the windspeed and surface
water inflow constant at 4.5 m/s and 10 kt/day, respectively. To model the
outflow we assume the outlet to be a pipe 0.2 $m^2$ in cross section and then
use Bernoulli's equation to model the water outflow based on the water level in
the cylinder above the pipe. We assume that the incoming heat follows a
combination of a third-order polynomial and a sinusoid function and that the
enthalpy of the incoming steam is 2.8 MJ/kg which corresponds to the enthalpy of
saturated steam at hydrostatic pressures similar to those at the bottom of the
lake \cite{Mayhew1978}. To the resulting synthetic observations we add
uncertainties similar to those estimated from real observations (see Section
\ref{realdata}).

@@

\begin{figure}
	\includegraphics[width=12 cm]{figures/synthetic_inversion_uks.png}  
    \caption{Inversion result for the synthetic test described in Section 
        \ref{syn_test}. Turquoise lines and markers represent hypothetical
        observations; red dashed lines are the inversion results and black
        dashed lines are the true input to the synthetic test.}
  \label{syn_example}
\end{figure}

\subsection{Real data}\label{realdata}
Temperature at RCL is measured by an integrated-circuit temperature device
(Texas Instruments LM35) located roughly 1.9 m below the lake's current
overflow level. The water level sensor is a hydrostatic pressure sensor (First
Sensor KTE/KTU/KTW6000\ldots CS Series) and it is co-located with the
temperature sensor. Both sensors have a sampling interval of 15 minutes and
transmit data several times a day via satellite to GNS Science, which monitors
New Zealand's geohazards through its GeoNet project (www.geonet.org.nz
 last accessed 16/07/2021).
Water samples are taken 8-10 times a year from which subsequently, among many
other analytes, concentrations of $Mg^{++}$ and $Cl^-$ are determined at GNS
Science's Geothermal Analytical Laboratory. We will focus here on $Mg^{++}$
concentration as it shows less deviation from the assumption of constant
dilution than $Cl^-$.

To estimate uncertainties of temperature and water level observations, we 
compute their daily averages and standard deviations. Uncertainties of $Mg^{++}$
concentrations are estimated directly when two samples were taken on the same day.
Otherwise, their uncertainty is taken to be the median of directly estimated
uncertainties.

$\dot{M_o}$ is measured once or twice per year and we further know from field
observations that the lake is not overflowing beneath a certain water level. We
fit a sigmoid function to these observations assuming a maximum outflow rate of
250 l/s. We assume the uncertainty in $\dot{M_o}$ as the 95 percent confidence
interval of the curve-fitting.

As there is no permanent weather station close to RCL we assume the wind speed
to be 5 m/s. This was the average windspeed derived from a temporary deployment
of a weather buoy on RCL and adjacent permanent weather stations
\cite{hurstUseWeatherBuoy2012}.

Figure \ref{inference_result} shows $\dot{Q_i}$ for RCL for lake measurements
between 4 March 2016 and 7 July 2021. During this time the heat input rate
reached a maximum of 753 MW and was on average 165 MW. The average standard
deviation was 34 MW and generally increased for lower values of $\dot{Q_i}$. To
put this in context, currently installed geothermal power plants in New Zealand
have a combined output of $\sim$~1000MW.

\begin{figure}
 	\includegraphics[width=13 cm]{figures/uks_real_data.png}  
 	\caption{Inference results for the heat input rate $\dot{Q_i}$ at RCL
             between 4 March 2016 and 7 July 2021. The solid line shows the
             marginal probability and the uncertainty is displayed as shaded
             region.}
\label{inference_result}
\end{figure}

\section{Discussion and conclusion}
Inferring the heat input rate into a crater lake is an important part in
estimating the total energy budget of a volcanic system \cite{Brown1989a}.
Similar to a volcano's gas budget, it can either indicate migration of magma or
changes in permeability within the vent system.

Figure \ref{psd} shows the power spectral density (PSD) for the marginal heat
input rate shown in Figure \ref{inference_result}. It was computed using the
multi taper method \cite{prietoComputersGeosciencesFortran2009} which provides
not only an estimate of the PSD but also its uncertainty. The PSD shows that the
dominant periods in the heat input rate lie between $\sim$100 and $\sim$250
days. Whether these periods are due to cyclic processes in the magmatic system,
the hydrothermal system, or a combination of both requires further modelling of
these processes. Many studies have demonstrated the value of non-linear Kalman
Filters for inference and data-assimilation with non-linear, high-dimension
numerical models \cite{White2018a}. While beyond the scope of this study, the
next logical step, therefore, is to include numerical models of magma degassing
and hydrothermal fluid flow to model the heat input rate, volume, and enthalpy
of the steam entering the lake.

\begin{figure}
    \includegraphics[width=12 cm]{figures/psd_analysis.png}  
 	\caption{Power spectral density of the heat input rate 
             (Figure \ref{inference_result}) shown as a solid line
             with uncertainty displayed as shaded region.}
    \label{psd}
\end{figure}

Irrespective of the process that is responsible for the periodicity, it appears
that it has not changed significantly over the past five years. This may
suggest that the volcano has been in a dynamic equilibrium during this time.
This is further supported by the cumulative heat input (Figure \ref{cumheat}).
It shows a slight increase in heat input in mid-2017 but a fairly constant rate
since then. Of course this does not in any way forecast the volcano's future
behaviour but it may serve as a useful baseline to detect changes in the
volcano's state more easily.

\begin{figure}
    \includegraphics[width=12 cm]{figures/cumulative_heat.png}
    \caption{Cumulative heat input into RCL (solid line) and its 
    uncertainty (shaded region).}
    \label{cumheat}
\end{figure}

Previous studies of RCL's mass and energy balance found that they required very
high values of enthalpy for the incoming steam to avoid violating physical
boundary conditions such as negative values for the inflow rate of meteoric
water \cite{hurstCraterLakeEnergy2015, Stevenson1992}. Their explanation was the
existence of a heat pipe beneath the lake where gas and fluid phases coexist. As
the rising gas enters the lake relatively cool lake water enters the vent
leading to an overall lower addition of mass to the lake and resulting in an
enthalpy value that is double than expected for saturated steam at hydrostatic
pressures likely to be found at the bottom of the lake. By including
uncertainties in our observations we no longer require these high enthalpy
values to satisfy the physical constraints. Numerical modelling of hydrothermal
processes may also shed further light on the heating processes and whether or
not a heat pipe is a plausible mechanism.

In conclusion, the combination of the mass and energy model, developed in
previous studies \cite{hurstCraterLakeEnergy2015, Stevenson1992}, with the
probabilistic inference method we developed here provides a powerful method to
continuously infer the heat input rate into RCL. It allows us to combine
disparate data streams, quantify our beliefs in the results and thereby augment
our monitoring capabilities by keeping a finger on Mt. Ruapehu's energy budget.

@@

where
\begin{table}[h!]
\begin{tabular}{cp{8cm}}
    A & is the surface of the lake (in $m^2$) \\
    $\sigma$ & is the Stefan-Boltzmann coefficient ($5.67e^{-8} \frac{W}{m^2 K^4}$) \\
    $\epsilon_w$ & is the lake emissivity \\
    $T_s$ & is the water surface temperature (here assumed to be 1 $^{\circ}C$ less than the
    measured temperature) \\
    $\epsilon_a$ & is the effective atmospheric emissivity \\
    $T_a$ & is the effective air temperature (here assumed to be 0.9 $^{\circ}C$) \\ 
\end{tabular}
\end{table}

\subsection{Evaporation}
As proposed by \cite{hurstCraterLakeEnergy2015} we use the equation of
\cite{adamsEvaporationHeatedWater1990} but replace the term for forced
convection by that of \cite{sartoriCriticalReviewEquations2000}:

@@

where
\begin{table}[h!]
\begin{tabular}{cp{8cm}}
    $T_{sv}$, $T_{av}$ & are the virtual surface and air temperatures respectively.
    This correction accounts for the extra buoyancy of water vapour compared to air
    (see Equation \ref{vtemp} for details) \\
    $e_s$, $e_a$ & are the saturation vapour pressure and the ambient air vapour
    pressure respectively \\
    $L$ & is the latent heat of evaporation \\
    $u$ & is the wind speed \\
    $X_0$ & is the fetch, or characteristic length, of the lake\\
    $P_a$ & is the atmospheric pressure \\
\end{tabular}
\end{table}

@@

\subsection{Sensible heat}
This describes the heat loss due to conduction and convection above a hot lake
and according to \cite{Stevenson1992} the ratio between sensible heat,
$Q_{sh}$, and heat loss due to evaporation, $Q_{evap}$ can be written as:

@@

\begin{table}[h!]
\begin{tabular}{cp{8cm}}
    $\rho$ & is the air density \\
    $c_a$ & is the specific heat of air\\
\end{tabular}
\end{table}
